A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-05 podc_paper/boundedspace.tex
working on description of GC

\section{Reducing Space}
\label{reducing}
\paragraph{Reducing Space Usage}
The \nf{blocks} arrays defined in our algorithm are unbounded. To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays, we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}. We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}). When a process wishes to write into location \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not. If not, it creates a shared array of size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}. Whether the \nf{CAS} is successful or not, \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array. When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps. The CAS Retry Problem does not happen here because if $n$ elements are appended to the array, then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}. Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to do the \nf{CAS} on \nf{arr[$i$]}. Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps. The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$. The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
\begin{figure}[hbt]  
  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
  \caption{Array segments.}
  \label{fig::doublingArray}
\end{figure}

@@

Tarjan \cite[Sec.~4.2]{Tar83} described a split algorithm for red-black trees that runs in logarithmic time.

>>>>

\section{Reducing Space Usage}
\label{reducing}

We now consider how to modify our implementation to bound the space it uses.
We replace the \fld{blocks} array in each node by a red-black tree (RBT)
that stores the blocks.
Each block has an additional \fld{index} field
that represents its position within the original \fld{blocks} array, and
blocks in a RBT are sorted by \fld{index}.
Accessing the block in entry $i$ in the \fld{blocks} array can then be 
replaced by searching the RBT for the block with index $i$.
The attempt to append a new block in entry $i$ of the \fld{blocks} array on line \ref{cas}
is replaced by an attempt to insert a new block with index $i$ into the RBT.

Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
However, we do not require all the standard search tree functions.
Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
concurrent \op{Refresh} succeeds in inserting a node, just as the CAS on line \ref{cas}
can fail if a concurrent \op{Refresh} does a successful CAS.
Thus, we can use a particularly simple concurrent RBT implementation.
A sequential RBT can be made persistent using the classic node-copying technique of 
Driscoll et al.~\cite{DSST89}:  all nodes are immutable, and operations on the 
tree are performed by making a new copy of each node $v$ that must be modified, as well
as each node along the path from the root to $v$.
The tree reachable from the new copy of the root is the result of applying the tree operation.
This only adds a constant factor to the running time of any routine designed for a (sequential) RBT.
Once a process has performed an update to the RBT representing the blocks of a node 
$v$ in the tournament tree, 
it uses a CAS to swing $v$'s pointer from the previous RBT root to the new RBT root.
A search in the RBT can simply read the pointer to the RBT root and perform a standard
sequential search on it.
Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way to implement a snapshot data structure.

To prevent RBTs from growing without bound, we would like to discard
blocks that are no longer needed.
Ensuring the size of the RBT is polynomial in $p$ and $q$ will 
also keep the running time of our operations polylogarithmic.
We msut keep a block $B$ if any element enqueued by the operations in $E(B)$ has
not yet been dequeued, since a later dequeue may need to traverse the block to 
find its response.
Because of the FIFO nature of queues, this means that we can throw away 
blocks whose indices are smaller than the block containing the oldest enqueue whose
value is still in the queue.
Fortunately, there is an efficient \op{Split} operation that can remove
all nodes with indices less than a given threshold from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.

To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
After every $p^2$ successful insertions into the RBT representing $\var{root}.\fld{blocks}$,
processes cooperate to perform \op{Split} operations on all nodes of the tournament tree  
to discard unneeded blocks.
We keep an array $\var{last}[1..p]$ where each process writes the index of the last block in
\var{root} that it dequeued an element from.
If a process performing a \op{Refresh} on the root finds that $root.\head$ is a multiple
of $p^2$, it does GC by first helping all pending dequeue operations by other processes,
then reading $\var{last}[1..p]$.

\here{need to complete the description of GC}

@@
